---
title: "Conditional Sampling in Mineral Rights Model"
output: html_notebook
author: "Stefan Heidekrüger"
date: 2021-04-28
---

In this notebook, we'll derive the conditional distribution $v_i|o_i$ in the 
generalizaiton of the "Mineral Rights" Model (Example 6.1 in Krishna, 2009).

Rather than drawing the common value standard-uniformly, we allow a slightly more
general 
$$v \sim U[lo, hi] $$ with arbitraty bounds $hi>lo\geq 0$.

Each players' individual observation is then given by $o_i = 2*v*x_i$ where $x_i$ are 
i.i.d. standard uniform variables. As such, observations take values in $[0,2v]$.

As the $x_i$ are independent, the $o_i|v$ are conditionally independent, and given
$v$, we can easily sample all players profiles. The main difficulty will therefore be
deriving a method to sample $v|o_i$. As this only concerns a single player,
we will drop all indices in the following.

Let's begin by doing some vizualisations

## Joint distribution of v and o.

```{r}
library(tidyverse)
theme_set(theme_minimal())
# let's use boundaries that are scaled and shifted, rather than 0,1

lo = 1
hi = 3

n = 2e6

v = runif(n, lo, hi)
x = runif(n)

o = 2*v*x

df = tibble(v=v, x=x, o=o)
df %>% head(200000) %>%  ggplot(aes(o,v)) + geom_bin2d() + scale_fill_viridis_c()
```

## Marginal distribution of $o$

With $v\sim U[lo,hi]$, we get that $o|v \sim U[0, 2v]$. As such we get the conditional density
$$f(o|v) = \frac{1}{2v}$$
on $[0,2v]$ and $0$ elsewhere.

We can than calculate the marginal pdf of $o$ via $f(o) = f(o|v)f(v)$. Note that the conditional density above implies
that $v\geq \frac o 2$ whenever $f(o|v)>0$. Any $v$ corresponding to an observation $o$ must therefore by larger than the "new low"
$$nl := max(lo, \frac o 2). $$

(We will use this term extensively, so let's give it a name.) With this, we get

$$f(o) = \int_{nl}^{hi} \frac{1}{2v} dv = \frac{log(hi) - log(nl)}{2(hi-lo)},$$
on $[nl, 2\cdot hi]$ and $0$ elsewhere. Note that this term only depends on $o$ via $nl$ (and the boundary).


```{r}
o_marginal_pdf = function(o) {
  if_else(0.5*o>lo,
          (log(hi)-log(0.5*o))/(2*(hi-lo)), 
          (log(hi) - log(lo))/(2*(hi-lo)))
  }
df %>% ggplot(aes(o)) + stat_density() + stat_function(fun=o_marginal_pdf, col='red', size=1)
```

## Conditional density of $v|o$

With $f(v), f(o), f(o|v)$ in place, we can calculate $f(v|o)$ via Bayes's thm:

$$f(v|o) = f(o|v)f(v)\frac{1}{f(o)} = \frac{1}{2v} \cdot \frac{1}{hi-lo}\cdot \frac{2(hi-lo)}{log(hi) - log(nl)} = \frac{1}{v \cdot (log(hi) - log(nl))}$$
restricted to $[nl, hi]$.

```{r}
my_o = 2.5
new_lo = pmax(my_o/ 2, lo)
c = log(hi) - log(new_lo)

pdf_v_given_o = function(v){1/(c*v)*(v>new_lo)*(v<hi)}


df %>% filter(my_o-0.02<o, o<my_o+0.02) %>% ggplot(aes(v)) + 
  stat_density(adjust=0.7, alpha=0.5) + 
  stat_function(fun=pdf_v_given_o, col='red') +
  ggtitle("Directly observed vs analytical conditional density of v|o for fixed/sliced o.")
```
## Sampling via icdf method

Given the pdf $f(v|o)$, we can derive its cdf

$$F(v|o) = \frac{1}{\log(hi) - \log(nl)} \int_{nl}^{hi} \frac{1}{v}dv = \frac{\log(v) - \log(nl)}{\log(hi) - \log(nl)}$$

Identifying the output with $u$ and inverting, we get the icdf

$$v|o = F^{-1}(u) = \exp\left(u\log(hi) - (1-u)\log(nl)\right) = hi^u \cdot nl^{1-u} $$
We can therefore sample $v|o$ via transforming a std-uniform RV $u$:

```{r}
u = runif(100000)
sampled_v_given_u = exp(u*log(hi) + (1-u)*log(new_lo))
tibble(v = sampled_v_given_u) %>% ggplot(aes(v)) + 
  stat_density(adjust=0.5) +
  stat_function(fun=pdf_v_given_o, col='red') +
  ggtitle("Observed kernel density estimate of v|o sampled via icdf vs analytical density.")
```

